---
title: 'Week 3 (LAB 1): Data Wrangling with dplyr'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author:
  - name: "Ali Abuzaid"
    affiliations:
      - name: "Spring 2025"
affiliation-title: "Quarter"
format: 
 pdf:

    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---

```{r setup-general, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)

bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```


```{r, include=FALSE}
library(tidyverse)
library(readr)
```


:::{.callout-tip}
# Submission Instructions


* This LAB must be completed and submitted **individually**. Collaboration is allowed for discussion, but each student must submit their own work.

* Ensure that all `R` code are presented clearly and appropriately.

* All figures should be numbered, and axes must be labeled. 

* Please use the provided `LAB 1.qmd` file to type your solutions and submit the completed LAB as a PDF file. You can utilize `RStudio` for this purpose. For guidance, refer to the [Tutorial: Hello, Quarto](https://quarto.org/docs/get-started/hello/rstudio.html)).

* Submit your solution via **Gradescope**.

:::
:::{.callout-caution}
# Due Date
**Due Date:** Thursday, April 17, 2025, 11:59 PM
:::
 
 \newpage
:::{.callout-important}
# **Exercises**

**Question 1**: 

Add a new column called `rank1` containing integers 1-4 expressing your personal taste preference (1 = tastiest). Make sure numbers are unique.

`r bfcolor("Replace this line with your answers", "red")` \



**Question 2**: 

Create a new dataframe `fruit_info_mod1` with an additional column `rank2` containing the same values as `rank1`.
`r bfcolor("Replace this line with your answers", "red")` \


**Question 3**: 

Remove all rank columns from `fruit_info_mod1` and store as `fruit_info_original`.

`r bfcolor("Replace this line with your answers", "red")` \


**Question 4**:

Rename columns to capitalize them using `rename()`.

`r bfcolor("Replace this line with your answers", "red")` \


**Question 5**: 

Check the dimensions of `baby_names` and store as `dimensions_baby_names`.


`r bfcolor("Replace this line with your answers", "red")` \


**Question 6**: 

Count how many names appear in each year and store as `occur_per_year`. Also store the number of distinct years as `num_years`.

`r bfcolor("Replace this line with your answers", "red")` \


**Question 7**: 

Find all entries for a friend's name and select just the Count, Sex, and Year columns.

`r bfcolor("Replace this line with your answers", "red")` \

**Question 8**: 

Find all girl names in 2010 that were given more than 3000 times.


`r bfcolor("Replace this line with your answers", "red")` \

**Question 9**: 

Find which name was most common in 1995 and how many babies had that name.
`r bfcolor("Replace this line with your answers", "red")` \


**Question 10**: 

Count how many distinct names there were for each sex in 1995.

`r bfcolor("Replace this line with your answers", "red")` \



::::
